---
format:
  html:
    toc: true
    toc-location: left
    df-print: kable
execute:
  echo: false
---

# Logging ROS1 medicational research

This logbook contains the logs of the research part of the bio-informatics minor at the Hanze in 2025-2026.

-   Author: Ramon Reilman
-   Class: BFV3
-   Minor: Bio-informatics

## First things First

The first thing done for this minor is researching the best working time for the medication series, we have no involvement in this part. We do need to select genes, done by Jasper Jonker, and design / find primers for these. This has to be done to QPCR (quantative) these given genes, so that we can do a transcriptomics analasis on the sequencing data.

### Gene choices
Literary research was done by Jasper, the findings and gene choices can be found in his [logbook](./Logbook_Jasper.html)

### Primers

After speaking to the head of the labwork, we came to an agreement on primer design. This agreement came wth a set of rules for designing these primers, listed below.

-   Melting temp: Preferably around 60 degrees celcius
-   Length range: 15-25 with a preference of 20 base pairs.
-   No circular primer forming.
-   Tools: primerblast

A list of genes will be created to be used for the qPCR. This list will be given by Jasper. I'm going to write a script that can download the sequences of DNA of these genes in fasta format.

```{r}
library(reticulate)
Sys.setenv(RETICULATE_AUTOCREATE = "FALSE")
use_condaenv("ROS1", required = TRUE)
py_config()
```

```{python}
from Bio import Entrez
```

```{python}
Entrez.email = "r.reilman@st.hanze.nl"
databases = Entrez.einfo()
print(Entrez.read(databases))
```

This contains a list of the different databases that can be accessed, we want to use the gene database to get the nucleotide sequences

```{python}
handle = Entrez.esearch(db="nucleotide", term="GRB2[gene] AND biomol_mrna[PROP] AND srcdb_refseq[PROP] AND human[ORGN]")
idlist = Entrez.read(handle)["IdList"]

handle2 = Entrez.efetch(db="nucleotide", id=idlist, rettype="gb", retmode="text")

```

```{python}
from Bio import SeqIO
seqs = ""
header = ""
for seqq in SeqIO.parse(handle2, "gb"):
    if "transcript variant 1" in seqq.description.lower():
        header = seqq.id + "_" + seqq.annotations["source"]
        seqs = seqq.seq
        break

```

This outputs many sequences for possibly the same gene, i want to check if they are the same sequences or not.

```{python}
header
```

All sequences are equal to eachother, so the first can be used. These can be turned into a multifasta format

```{python}
fasta_showcase = f">{header}\n{seqs}"
fasta_showcase
```

```{python}
with open("/students/2025-2026/ros1_transcriptomics/minor_ROS1/data/showcase.fasta", "w", encoding='utf-8') as f:
    f.write(fasta_showcase)
    
```

And just like that we have a fasta file containing the sequence of a given gene. I will now create a script that takes a file of gene names, and turns in into a multifasta file.

These fasta files will be pulled as the mRNA variant 1 transcripts of the genes, the GRB2 variant 1 page can be found [here](https://www.ncbi.nlm.nih.gov/nuccore/NM_002086.5)

We can design primers for these sequences using primer blast. Which takes some parameters to create these primers. I'm going to do a test-run with the fasta file generated above. I will load in these primers, with the selected parameters and discus the output of primer blast.

```{r}
packages <- c("config")
install.packages(setdiff(packages, rownames(installed.packages())))
```

```{r}
library(yaml)
config <- read_yaml("/students/2025-2026/ros1_transcriptomics/minor_ROS1/configs/config.yaml")
```

```{r}
primer_data <- read.csv2(config$primer_data, sep = "\t")
primer_data
```

This dataframe contains the ouput of a primerblast run, i want to explore the different columns, and see what they mean.

The first couple of columns are about the **forward primer** starting with basic information about the length of the primer and where the primer starts/ends. The more interesting metrics show up after this:

- Tm, the primer melting temperature
- Self complementary, a metric to display how much a primer can bind to itself (or copies of itself) this is naturally not wanted, the lower the value is, the better.
- 3\` self complementary, similar to the self complementary metric, but now specifically for the 3\` end. Naturally unwanted too, so the lower the better.

The same metrics exist for the **reverse primer**

## Transcriptomics

We've gotten data from different research to prepare our pipeline for the sequencing data that will arrive in a couple of weeks. This data has been downloaded.

Jasper and I are thinking of creating a snakemake pipeline to process our data, considering this; We will create a complete pipeline for transcriptomics, starting with the downloading of data - turning it into fasta files and even creating an indexed reference genome. All of these steps, and more will be found in this (and Jasper's) logs.

The first thing I want to do while we are waiting for the downloading of data is creating an indexed reference genome using STAR. We've done this before, so should not be that much of an effort.

```{bash, eval=FALSE}
sbatch --partition=assemblix --job-name=hg38_index --ntasks=1 --cpus-per-task=32 --mem=300G --time=10:00:00 --wrap="/students/2025-2026/ros1_transcriptomics/bin/STAR_2.7.11b/Linux_x86_64/STAR --runThreadN 4 --runMode genomeGenerate --genomeDir /students/2025-2026/ros1_transcriptomics/genomes/hg38/index --genomeFastaFiles /students/2025-2026/ros1_transcriptomics/genomes/hg38/GCA_000001405.29_GRCh38.p14_genomic.fna --sjdbGTFfile /students/2025-2026/ros1_transcriptomics/genomes/hg38/GCA_000001405.29_GRCh38.p14_genomic.gtf --sjdbOverhang 75"
```

```{bash, eval=FALSE}
sbatch --partition=assemblix --job-name=hg19_index --ntasks=1 --cpus-per-task=32 --mem=300G --time=10:00:00 --wrap="/students/2025-2026/ros1_transcriptomics/bin/STAR_2.7.11b/Linux_x86_64/STAR --runThreadN 4 --runMode genomeGenerate --genomeDir /students/2025-2026/ros1_transcriptomics/genomes/hg19/index --genomeFastaFiles /students/2025-2026/ros1_transcriptomics/genomes/hg19/GCA_000001405.14_GRCh37.p13_genomic.fna --sjdbGTFfile /students/2025-2026/ros1_transcriptomics/genomes/hg19/GCA_000001405.14_GRCh37.p13_genomic.gtf --sjdbOverhang 75"
```

Both the test data has been downloaded, and the genomes have been indexed. This means i will have to fasta-dump the sra files, and then afterwards map these.

```{bash, eval=FALSE}
find /students/2025-2026/ros1_transcriptomics/GSE214715/SRA -name "*.sra" | \
  parallel fasterq-dump -O /students/2025-2026/ros1_transcriptomics/GSE214715j/fasta/ {} &
```

```{bash, eval=FALSE}
find /students/2025-2026/ros1_transcriptomics/GSE239844/SRA -name "*.sra" | \
  parallel fasterq-dump -O /students/2025-2026/ros1_transcriptomics/GSE239844/fasta/ {} &
```

We can continue with the mapping process after this dumping is done. We will sure star for this. I will do a singular test run on 2 fasta files, then afterwards i will implement STAR into snakemake and run it via that, it will be run in parallel, and will make a nice start to our pipeline. This pipeline will eventually be extended by a bit of quality control, but mostly R scripts to generate different plots and statistical results.

```{bash, eval=FALSE}

STAR --runThreadN 6 \
  --genomeDir /students/2025-2026/ros1_transcriptomics/genomes/hg38/index/ \
  --readFilesIn /students/2025-2026/ros1_transcriptomics/GSE214715/fasta/SRR21789608_1.fastq /students/2025-2026/ros1_transcriptomics/GSE214715/fasta/SRR21789608_2.fastq \
  --outSAMtype BAM SortedByCoordinate  \
  --quantMode GeneCounts  \
  --genomeLoad LoadAndRemove \
  --limitBAMsortRAM 20000000000 \
  --outFileNamePrefix /students/2025-2026/ros1_transcriptomics/GSE214715/mapping/test_hg38_star_
  

```

```{bash, eval=FALSE}

STAR --runThreadN 8 \
  --genomeDir /students/2025-2026/ros1_transcriptomics/genomes/hg38/index/ \
  --readFilesIn /students/2025-2026/ros1_transcriptomics/GSE239844/fasta/SRR25486700_1.fastq /students/2025-2026/ros1_transcriptomics/GSE239844/fasta/SRR25486700_2.fastq \
  --outSAMtype BAM SortedByCoordinate  \
  --quantMode GeneCounts  \
  --genomeLoad NoSharedMemory \
  --outFileNamePrefix /students/2025-2026/ros1_transcriptomics/GSE239844/mapping/hg38/SRR25486700
  

```

The results of STAR will end up as multiple read count files, i will create a script (functions defined below first) that will process these different counts, and replace the ensembl gene names with human readable gene names.

```{r}
config$src_dir

```

```{r}
library(glue)

read_sample <- function(file.name) {
  sample.name <- strsplit(file.name, ".", fixed = TRUE)[[1]][1]
  sample <- read.table(file.name, header = FALSE, sep="\t", 
                       row.names = NULL, skip = 4)
  names(sample)[2] <- sample.name

  return(sample[c(1, 2)])
}

merge_counts <- function(genome, exp) {
  setwd(glue("{src}/{exp}/mapping/{genome}/", src = config$src_dir, exp = exp, genome = genome))
  file.names <- list.files('./', pattern = "*ReadsPerGene.out.tab")
  counts <- read_sample(file.names[1])
  
  for (file.name in file.names[2:length(file.names)]) {
  sample <- read_sample(file.name)
  counts <- merge(counts, sample, by = 1)
  }
  rownames(counts) <- counts$V1
  counts <- counts[-1]
  return(counts)
  
}


```
```{r}
genomes = c("hg19", "hg38")
experiments = c("GSE214715", "GSE239844")
for (exp in experiments) {
  for (genome in genomes) {
    counts <- merge_counts(genome, exp)
  }
}

```

The gene names resulting from this are Ensembl transcripts, i will use biomaRt to transcribe these and annotate these to readable gene names.
```{r}

library(biomaRt)

```

```{r}
df <- data.frame(counts)

```

```{r}
ensembl_ids <- gsub("\\..*$", "", rownames(df))

ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

gene_info <- getBM(
    attributes = c("ensembl_gene_id", "hgnc_symbol", "external_gene_name"),
    filters = "ensembl_gene_id",
    values = ensembl_ids,
    mart = ensembl
)


```

```{r}
duplicates <- gene_info[duplicated(gene_info$ensembl_gene_id) | 
                        duplicated(gene_info$ensembl_gene_id, fromLast = TRUE), ]

# Sort to see them together
duplicates <- duplicates[order(duplicates$ensembl_gene_id), ]
head(duplicates, 10)
```

We see here that there are 2 duplicates happening, i will just use the unique version of the hgnc_symbol

```{r}
library(dplyr)

gene_info_unique <- gene_info %>%
  group_by(ensembl_gene_id) %>%
  summarise(
    hgnc_symbol = paste(unique(hgnc_symbol[!is.na(hgnc_symbol) & hgnc_symbol != ""]), 
                       collapse = "/"),
    external_gene_name = first(na.omit(external_gene_name))
  ) %>%
  ungroup()


merge_counts_with_symbols <- function(counts, gene_info_unique) {
  ensembl_ids <- gsub("\\..*$", "", rownames(counts))
  
  counts_df <- data.frame(
    ensembl_gene_id = ensembl_ids,
    counts,
    row.names = NULL
  )
  
  merged_df <- merge(
    counts_df,
    gene_info_unique[, c("ensembl_gene_id", "hgnc_symbol")],
    by = "ensembl_gene_id",
    all.x = TRUE
  )
  
  merged_df$hgnc_symbol[is.na(merged_df$hgnc_symbol)] <- 
    merged_df$ensembl_gene_id[is.na(merged_df$hgnc_symbol)]
  
  merged_df$hgnc_symbol[merged_df$hgnc_symbol == ""] <- 
    merged_df$ensembl_gene_id[merged_df$hgnc_symbol == ""]
  
  rownames(merged_df) <- make.unique(merged_df$hgnc_symbol, sep = "_")
  
  final_counts <- merged_df[, !names(merged_df) %in% c("ensembl_gene_id", "hgnc_symbol")]
  
  return(final_counts)
}

```

```{r}
counts <- merge_counts_with_symbols(df, gene_info_unique)
```

Now the gene symbols are set as the row names, i will turn this functionality in an R script so that it can be used in the snakemake pipeline.

## DESEQ2

So now that we have the counts data ready, merged and annotated, we can start working on the actual analysis. Both me and Jasper will work on the experiments respectively, which means that I will be working on GSE23. The analyses will be done using deseq2, as we both have done before. Our plan for this is to create different functionalities that can be used in the snakemake pipeline, so that when we eventually get our data, we will be able to generate results rather fast in 1 go.

```{r}
library(DESeq2)

```


```{r}

data <- read.csv2(paste0(config$counts_dir, "/hg38_counts.csv"), sep = ",", row.names = 1)
col_data <- read.csv2(config$col_data, sep = ";")
rownames(col_data) <- col_data[,1]
col_data$treatment <- factor(col_data$treatment, levels = c("DMSO", "250 nM crizotinib"))
head(data)
```

This is the count data for the GSE23 experiment, mapped to the hg38 human genome. There is also another file being loaded in; The column data file.

```{r}
head(col_data)
```

This contains information about the different samples. To select certain samples or groups of samples i will generate masks for all possible assay and treatment combination. First i will make sure that the order of the columns of the count data is the same as the Run column in the col_data


```{r}
data <- data[, rownames(col_data)]
all(rownames(col_data) == colnames(data))
```

This means that all of the rownames of the coldata is alligned with the colname of the counts data. Which means I can now generate masks for the different groups, i will also rename the SRR reads so that they're named after the assay type and treatment

```{r}
col_data <- col_data %>%
  dplyr::group_by(Assay.Type, treatment) %>%
  dplyr::mutate(r_num = row_number()) %>%
  dplyr::ungroup() %>%
  mutate(condition = paste0(Assay.Type, "_", treatment, "_r", r_num))

head(col_data)

```

This will make it more readable what every point is, including which one of the tripples it is.

```{r}
for (i in 1:nrow(col_data)) {
  idx <- grep(col_data$Run[i], names(data))
  names(data)[idx] <- col_data$condition[i]
}
```


```{r}
head(data)

```

Now the names of the SRR files align with more readable names, we can now seperate them into groups

```{r}
HCC78_crizo <- grep("HCC78_250 nM crizotinib", names(data))
HCC78_DMSO <- grep("HCC78_DMSO", names(data))

CUTO38_crizo <- grep("CUTO38_250 nM crizotinib", names(data))
CUTO38_DMSO <- grep("CUTO38_DMSO", names(data))

CUTO37_crizo <- grep("CUTO37_250 nM crizotinib", names(data))
CUTO37_DMSO <- grep("CUTO37_DMSO", names(data))

CUTO33_crizo <- grep("CUTO33_250 nM crizotinib", names(data))
CUTO33_DMSO <- grep("CUTO33_DMSO", names(data))

CUTO28_crizo <- grep("CUTO28_250 nM crizotinib", names(data))
CUTO28_DMSO <- grep("CUTO28_DMSO", names(data))

CUTO27_crizo <- grep("CUTO27_250 nM crizotinib", names(data))
CUTO27_DMSO <- grep("CUTO27_DMSO", names(data))

CUTO23_crizo <- grep("CUTO23_250 nM crizotinib", names(data))
CUTO23_DMSO <- grep("CUTO23_DMSO", names(data))
```

Now we can start with DESEQ, finally.

```{r deseq-analysis, cache=TRUE}
dds <- DESeqDataSetFromMatrix(countData = data,
                              colData = col_data,
                              design = ~ Assay.Type + treatment)

dds <- DESeq(dds)
```

With this we can do a Differential Expression Analysis

### Differential Expression Analysis


```{r}
res <- results(dds, contrast=c("treatment", "250 nM crizotinib", "DMSO"))

res

```

This is the result of comparing the treatments (either no treatment or 250 nM of crizotinib). I need to shrink the log fold change, this is usefull for the visualisations.

```{r}

print(resultsNames(dds))
resLFC <- lfcShrink(dds, coef = "treatment_250.nM.crizotinib_vs_DMSO", type = "apeglm")
```


```{r}
vsd <- vst(dds, blind = F)
pcaData <- plotPCA(vsd, intgroup= c("Assay.Type", "treatment"), returnData=T)

```

I want to generate a PCA plot to look at the different samples, and see if they align or if there are any outliers. I will do this using the given PCA data from deseq2 and generating a PCA plot from this using ggplot.

```{r}
library(ggplot2)
```
Before I do that, i will take a peak at the given PCA data:
```{r}
head(pcaData)
```

Here we see a couple of different interesting things:
- The 2 PCA values used for plotting
- It also stores the treatment, assay-types and triplo-number.

```{r}
percentVar <- round(100 * attr(pcaData, "percentVar"))

ggplot(pcaData, aes(PC1, PC2, color = Assay.Type, shape = treatment)) +
  geom_point(size = 3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance"))

```

Interesting is that some of the cell types like CUTO23 and 33 don't seem to respond to the treatments, as they are all clustered very tightly around 1 area. We see a clearer seperation with CUTO 38 and CUTO 27. 